Gradual Soundness: Lessons from Static Python

### 2021-12-30 pj-7.1/paper.tex
pj: more in-comment notes from carl

@@

%% intro sentence
The dynamic type is also different than the flexible dynamic type provided
by true gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
for a standard type precision relation.
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{In addition, C types~(\cref{s:ctypes}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.
  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors are a consequence of \SP{} being driven by the needs of an
existing codebase.
The literature on gradual types has served only as a guide, not a mandate.

@@

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:ctypes}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and false get enforced with a run-time check
(\code{Union[T, False]}, or \code{Optional[t0]}).

@@

\subsubsection{Unboxed Primitive}
\label{s:ctypes}

@@

@@

@@

@@

%% - pushback one time about removing MI
%% - otherwise, people happy with low-cost improvements

@@

@@

Optimize Python integer operations (in addition to the ctypes).
Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.

>>>>

%% 2021-12-23: TODO
%% Carl: The behavior of dynamic type in static code should be:
%%   a) any type can flow to dynamic
%%   b) dynamic can flow to any type, but it requires the insertion of a runtime cast for soundness
%%   c) dynamic type should emit the same bytecode for any given operation that non static code would emit
%%      ... I think differing behavior around (c) is also unlikely — the structure of our
%%      code generation is that we call methods on type objects to emit byte code for
%%      various operations on a value of that type, and our dynamic type simply
%%      forwards all those calls to the default Python compiler, which should always
%%      result in dynamic type emitting the same byte code that nonstatci code would.
%%      This isn’t universal, though (i.e. for some operations, e.g. binary ops, we
%%      might explicitly handle various cases for operand type pairings, not always by
%%      calling a polymorphic method on each operand type), which could open the door
%%      to differing behavior here. But I can’t think of or find any such intentional
%%      cases, and I suspect if any did crop up we’d consider it a bug.

@@

%% intro sentence
The dynamic type is also different than the flexible dynamic type provided
by true gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
for a standard type precision relation.
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.

    %% 2021-12-13: TODO NOTE SWALLOW MORE
    %% I think the fundamental unusual thing we do here is our per-module
    %% incremental adoption strategy for our compiler. We don’t apply our compiler
    %% to the entire codebase, using presence or absence of type annotations as
    %% the sole indicator of whether to compile code with static or dynamic
    %% behavior. We apply our compiler only to selected opted-in modules, and the
    %% rest of the codebase is compiled with Python’s default compiler and no
    %% type-checking (it may of course be checked by Pyre but that’s unrelated to
    %% compilation.) So this clearly opens the door to differing behavior between
    %% code that is compiled by our compiler and code that isn’t, even when the
    %% former has some or many values/expressions typed as dynamic. It also makes
    %% such differences somewhat more acceptable in practice — since adoption of
    %% an individual module into our compiler is opt-in, we can swallow more
    %% behavior difference implied solely by that change.

  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors are a consequence of \SP{} being driven by the needs of an
existing codebase.
The literature on gradual types has served only as a guide, not a mandate.

@@

%% TODO carl 2021-12-24 ... very important to introduce with ChkDict ... GG not necessary only nice-to-have
%% > One area where we definitely violate the gradual guarantee and always will
%% > is primitive machine types. Unlike Python object types these do not flow to
%% > dynamic, and dynamic can never flow to them! Effectively they form a
%% > fully-typed subset of the type system that does not participate in gradual
%% > typing and can never occur in non-static code. So this violates GG because
%% > there are many examples where removing an annotation of a primitive machine
%% > type can cause static code to no longer type check.
%% > 
%% > We are ok with this violation of the gradual guarantee because the form of
%% > the gradual guarantee we really care about is not “removing an arbitrary
%% > annotation” but rather “an untyped module compiles without changes when
%% > adopted into the SP compiler” — and an untyped module cannot have
%% > pre-existing uses of the primitive machine types, since those types cannot
%% > exist outside SP.

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
(\code{Union[T, None]}, or \code{Optional[t0]}).

@@

\subsubsection{Unboxed Primitive}
\label{s:c-types}

%% 2021-12-29: TODO
%% Using the term “ctypes” for these might be confusing, since the Python stdlib
%% has a “ctypes” module that isn’t related to our primitive types. We do call the
%% base type for them CType in our compiler, but when discussing them we usually
%% call them “primitive types” or “machine types” — “C types” could also work but
%% I wouldn’t combine it into one word :)
%%
%% I just did a very rough grep check and found that we have only 12 modules
%% currently using primitive types: all int32 and int64 and cbool. So they are
%% used pretty sparingly, but they are also definitely critical to our ability to
%% match Cython in a few of the  hottest code paths that used to be Cython.

@@

%% TODO 2021-12-28
%% - model soundness argument = similar to Nom, Redex tests
%% - while True break another example of modeling helping out
%%   it's a second implementation, helpful to find bugs
%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us

@@

%% TODO 2021-12-29: Carl
%% "full soundness" requires strict modules, means you can trust module toplevel
%% hmph

@@

%% TODO 2021-12-29 Carl
%% We definitely did find some regressions (especially when converting Cython, but
%% also one or two in non-Cython conversions.) But we were never satisfied with a
%% regression, and we were always able to apply the JIT or improve SP
%% optimizations or adjust the code to turn the regression neutral or positive.
%% (In fact I think all? the non-Cython regressions I can recall boiled down to
%% “we forgot to enable the JIT for an important function” and just rectifying
%% that took care of it.)

%% TODO think about it 2021-12-29 Ben
%% > + Is there any sense in which the 530 converted modules are a big
%% > sample? Maybe not, since as you say 530 is tiny for the codebase as a
%% > whole. But do they exercise all sorts of behavior? Do they form a big
%% > sample of certain parts of the Instagram codebase?

@@

%% - pushback one time about removing MI multiple inheritance
%% - otherwise, people happy with low-cost improvements

@@

FILL

@@

Optimize Python integer operations (in addition to the c types).
Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
